# Variables {#variables}

Before reading the chapter, please download the [exercise notebook](notebooks/Seminar 02 - Vectors.Rmd), put it into your seminar project folder and open the project. You need both the text and the notebook with exercises to be open, as you will be switching between them.

## Vectors and Scalars (and Lists, which are a special case) {#vectors}

Before we can start programming in R, you need to learn about vectors. This is a key concept in R, so your understanding of it will determine how easy it will be for you to use R. Do all of the exercises, even if they feel simple and repetitive, and do not hesitate to ask me whenever something is unclear. Remember, you need to master vectors before you can master R!

Often, a concept of a variable is describe as a box you can put something in. A box has a name tag on it, which is the _name_ of the variable. Whatever you put in is the _value_ that you store. 

**Figure with a the box**

This "putting in" concepts is reflected in R syntax
```{r eval=FALSE}
number_of_participants <- 10
```
Here, `number_of_participants` is the name of the variable (name tag for the box we will be using), `10` is the value you store, and `<-` means **"put `10` into variable `number_of_participants`"**. If you know other programming languages, you probably expected the usual assignment operator `=`. Confusingly, you can use it as well, but there are some subtle, yet important, differences in how they operate behind the scenes. We will meet `=` again when we will be talking about functions and, in particular, Tidyverse way of doing things but for now **only use `<-` operator**! 

Unfortunately, the box metaphor you've just learned, doesn't quite work for R. Historically, R was developed as a language for statistical computing, so it was  based on concepts of linear algebra instead of being a "normal" programming language. This means that the primary data unit in R is a **vector**, which you may remember from geometry or, hopefully, from linear algebra, as an arrow that goes from 0 to a specific point in space. But from computer science point of view, a vector is just a list of numbers (or some other values, as you will learn later). This means that there are no "single values" in R per se, there are only vectors variable length including vectors of length 1,  which are usually called _scalars_ ^[Multiplication of a vector by another vector _transforms_ it but for a single element vector the only transformation you can get is "scaling", hence, the name.]. With respect to the "box metaphor", this means that we have a box with numbered (indexed) slots in it. A simple assignment makes sure that "the box" has as many slots as the values you want to put in and stores these values one after another starting with slot #1. So, the example above `number_of_participants <- 10` creates a variable with one (1) slot and stores the value in it.

**Figure with a slot in the box**

But the single value (vector with length of one) is a special case. More generally you write:
```{r eval=FALSE}
response <- c(1, 7, 3)
```
Here, you create a variable (box) named `response` that has three slots in it because you want to store three values. You put values `1, 7, 3` into the slots #1, #2, and #3. The `c(1, 7, 3)` notation is how you create a vector in R by [**c**oncatenating](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/c) (or **c**ombining) values^[I find this to be a poor choice of name but we are stuck with, so just get used to it.]. The figure below illustrates the idea:

**Figure with a the 3 slots in box**

Building on the box metaphor: If you can store something in a box, you can take it out! In the world of computers it works even better, because rather than taking something out, you just make a copy of that and store this copy somewhere else or to use it to compute things. Minimally, we would like to see what is inside of the box. For this, you can use [print](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/print) function:

```{r}
response <- c(1, 7, 3)
print(response)
```

Or, we can make a copy of values in one variable and store them in another:
```{r}
x <- c(3, 6, 9)
y <- x 

print(x)
print(y)
```
First, we  create vector of length 3 via concatenation `c(3, 6, 9)`, then create a 3-slot variable `x` and put the numbers into that variable. Next, we make a copy of these three values and store them in a different variable `y`. Importantly, the values in varible `x` stayed as they were. Take a look at the figure below, which graphically illustrate this:

**Figure on copying values box**

**Do exercise #1.**

Remember, everything is a vector! This means that `c(3, 6, 9)` does not **c**oncatenate numbers, it **c**oncatenates three length one vectors (scalars) `3`, `6`, `9`. Thus, **c**oncatenation works on longer vectors in exactly the same way:
```{r}
x <- c(1, 2, 3)
y <- c(4, 5)
print(c(x, y))
```
**Do exercise #2.**

Everything is a vector! This means that when you are using a function, you are always applying (mapping) it to a vector. This, in turn, means that you can apply the function to all values in one go. For example, you can compute a cosine of all values in the vector.
```{r}
cos(c(0, pi/4, pi/2, pi, -pi))
```
In contrast, in Python or C, you would need to loop over the values and compute the cosine for one value at a time (matrix-based NumPy library is a different story). Or think about Excel, where you need to extend formula over the rows but each row is computed independently (so you can deliberately or accidentally miss some rows). In R, because everything is the vector, the function is applied to every value automatically. Similarly, if you are using aggregating functions, such as `mean()` and `max()`, you can pass a vector and it will return a length-one vector with the value.

```{r}
mean(c(1, 2, 6, 9))
```
## Vector indexing, single value (#vector-index-single)
A vector is an ordered list of one or more values (box with one or more slots) and, sometimes, you need only one of the values. Each value (slot in the box) has its own index from 1 till N, where N is the length of the vector. To access that slot you use square brackets `some_vector[index]`. You can both get or set the value for the individual slots the same way as for the whole vector.

```{r}
x <- c(1, 2, 3)
# set SECOND element to 4
x[2] <- 4

# print the entire vector
print(x)

# print only the third element
print(x[3])
```
If your vectors contains five values, you would expect that using indexes below `1` or above `5` would generate an error. However, in R things are bit more complicated. Index of `0` is a special case and produces an _empty vector_ (vector of length zero).
```{r}
x <- c(1, 2, 3)
x[0]
```

If you try to get vector element using index that is larger than vector length (so `6` if your vector length is 5), R will return `NA` ("Not Available" / Missing Value). 

```{r}
x <- c(1, 2, 3)
x[5]
```
In both cases, it won't throw an error or even warn you!

When **setting** the value by index, using `0` will produce no effect (and, also, no error or warning).
```{r}
x <- c(1, 2, 3)
x[0] <- 5
print(x)
```
If set an element those index is **larger** than vector length, the vector will be automatically expanded to that length and all the elements between the old values and the new one will be `NA`.
```{r}
x <- c(1, 2, 3)
x[10] <- 5
print(x)
```
I am bringing these details up to serve as a general warning about R. It is a very powerful language that allows you to be very flexible. Unfortunately, that flexibility means that base R won't stop you from shooting your foot or, even worse, you won't even notice that your foot is shot because it won't generate either errors or warnings, as in example above. Good news id that things are far more restricted and consistent in Tidyverse.

**Do exercise #5.**

Finally, you can use _negative_ indexes. In that case, you **exclude** the value with that index and return or modify the rest.
```{r}
x <- c(1, 2, 3, 4, 5)
# this will return all elements but #3
x[-3] 
```
```{r}
x <- c(1, 2, 3, 4, 5)
# this will assign new value to all elements but #2
x[-2] <- 10
x
```
Given that negative indexing returns everything **but** the indexed value, what do you think will happen here (**Do exercise #6**)?
```{r eval=FALSE}
x <- c(10, 20, 30, 40, 50)
x[-10]
```
## Slicing  {#vector-index-slicing}
So far we were reading or modifying either the whole vector or just one element of it. But the index you pass in square brackets (you've guess it!) is also a vector! Which means that you can construct a vector of indexes the same way you construct a vector of any values (the only restriction is that index values must integers).
 
```{r}
x <- c(10, 20, 30, 40, 50)
x[c(2, 3, 5)]
```
Moreover, when constructing a vector index, you can put the index values in the order you require (starting from the end of it, random order, etc.) or use the same index more than once.
```{r}
x <- c(10, 20, 30, 40, 50)
x[c(3, 5, 1, 1, 4)]
```
You can also use several negative indexes to exclude multiple values and return the rest. Here, neither order nor the duplicate indexes matter: Regradless of which value you exclude (by index) the first or how many times you exclude it, you still get _the rest_ of the vector in its default order.
```{r}
x <- c(10, 20, 30, 40, 50)
x[c(-4, -2, -2)]
```
Note that you cannot mix positive and negative indexes as R, finally, will generate an error.
```{r eval=FALSE}
x <- c(10, 20, 30, 40, 50)

# THIS WILL GENERATE AN ERROR: 
# "Error in x[c(-4, 2, -2)] : only 0's may be mixed with negative subscripts"
x[c(-4, 2, -2)]
```
Finally, here, zero index makes no difference.
```{r}
x <- c(10, 20, 30, 40, 50)
x[c(1, 0, 5, 0, 0, 2, 2)]
```
**Do exercise #X.**

## Colon Operator and Sequence Generation {#colon-sequence}
To simply vector indexing, R provides you with a shortcut to create a range of index. An expression `A:B` (a.k.a.[Colon Operator](https://stat.ethz.ch/R-manual/R-devel/library/base/html/Colon.html)) builds a sequence of integers starting with `A` and ending with **and including(!)** `B` (the latter is not obvious, if you come from Python).
```{r}
3:7
```
Thus, you can use it to easily create an index and, because everything is a vector!, combine it with other values.
```{r}
x <- c(10, 20, 30, 40, 50)
x[c(1, 3:5)]
```
The sequence above is increasing but you can also use the colon operator to construct a decreasing one.
```{r}
x <- c(10, 20, 30, 40, 50)
x[c(5:2)]
```
The colon operator is limited to sequence with steps of `1` or `-1`, so for more flexibility you can use `seq(from, to, by, length.out)` (a.k.a [Sequence Generation](https://stat.ethz.ch/R-manual/R-devel/library/base/html/seq.html)). The `from` and `to` are starting and ending values (just like in the colon operator) and you can specify either a step via `by` parameter (as, in "from A to B by C") or via `length.out` parameter (how many values you want to generate).
Using `by` version:
```{r}
seq(1, 5, by=2)
```
Same sequence but using `length.out` version:
```{r}
seq(1, 5, length.out=3)
```
You have probably spotted the `=` symbol. Here, it is not an assignment but is used to specify the parameters when you call a function. Thus, we are still sticking with `<-` **outside** of the function calls but are using `=` **inside** the function calls.

**Go to exercise #X**.


## Working with two vectors

You can also use mathematical operations on several vectors. The vectors are matched element-wise. Thus, if you add two vectors of **equal** length, the _first_ element of the first vector is added to the _first_ element of the second vector, _second_ element to _second_, etc.
```{r}
x <- c(1, 4, 5)
y <- c(2, 7, -3)
print(x + y)
```

**Do exercise to #3.**

What if vectors are of different length? If the length of the longer vector is a multiple of the shorter vector length, the shorter vector is repeated N-times (where $N = length(longer~vector) / length(shorter~vector)$) and this repeated vector is then used for the mathematical operation. Take a look at the results of the following computation

```{r}
x <- c(1, 1, 1, 1, 1, 1)
y <- c(2, 3)
print(x + y)
```
Here, the values of `y` were repeated three times to match the length of `x`, so the actual computation was `c(1, 1, 1, 1, 1, 1) + c(2, 3, 2, 3, 2, 3)`. A vector of length 1 (scalar) is a special case because any integer is a multiple of 1, so that single value is repeated `length(longer_vector)` times before the operation is performed.

```{r}
x <- c(1, 2, 3, 4, 5, 6)
y <- 2
print(x + y)
```
Again, the actual computation is `c(1, 2, 3, 4, 5, 6) + c(2, 2, 2, 2, 2, 2)`. 

Finally, if the length of the longer vector **is not** a multiple of the shorter vector length, R will repeat the shorter vector N times, so that $N = ceiling(length(longer~vector) / length(shorter~vector))$ (where [ceiling()](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/Round) rounds a number up) and truncates (throws away) the extra elements it does not need. Although R will do it, it will also throw a warning about mismatching objects' (vectors') lengths.

```{r}
x <- c(2, 3)
y <- c(1, 1, 1, 1, 1)
print(x + y)
```
One thing to keep in mind: R does this length-matching-by-repetition automatically and shows a warning only if two lengths are not multiples of each other. This means that vectors will be matched by length even if that was not your plan (one vector is half the length because you accidentally forgot to add data from second experimental block) and you do not realize that. Thus, do make sure that your vectors are matched in their length, so that you are not caught surprised by this behavior (you can use function [length()](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/length) for this).

**Do exercise #4.**

## Wrap up
By now you have learned more about vectors and vector operations in R than you probably bargained for. Admittedly, not the most exciting topic! And there was not a single word on psychology, data analysis, or statistics. However, R is obsessed with vectors (everything is a vector!) and understanding them will make it easier to understand tables (lists of vectors, more on that next time) and functional programming, another two core concepts of R. Finish this seminar off by doing the remaining exercises. Let's see whether R can still surprise you!

