# Missing data

## Missing data
Sometimes data is missing. It can be missing _explicitly_ with `NA` standing for [Not Available / Missing](https://stat.ethz.ch/R-manual/R-devel/library/base/html/NA.html) data. Or, it can be missing _implicitly_ when there is no entry for a particular condition. In the latter case, the strategy is to make missing values explicit first (discussed below). When you have missing values, represented by `NA` in R, you must decide how to deal with them: you can use this information directly as missing data can be diagnostic in itself, you can impute it using either a sophisticated statistical methods or via a simple average/default value strategy, or you can exclude them from the analysis.

## Completing data {#complete}
To make implicit missing data explicit, _tidyr_ provides function [complete()](https://tidyr.tidyverse.org/reference/complete.html) that figures out all combinations of values for columns that you specified, finds missing combinations and adds with using `NA` or a specified default value. Imagine our toy table was incomplete (no data for Participant `2` and Face `M-2`)
```{r echo=FALSE}
incomplete_df <- head(widish_df, 3)
```
```{r echo=FALSE}
incomplete_df %>%
knitr::kable()
```

We can [complete](https://tidyr.tidyverse.org/reference/complete.html) that table by specifying columns that define all required combinations.
```{r}
complete_df <- complete(incomplete_df, Participant, Face)
```
```{r echo=FALSE}
complete_df %>%
knitr::kable()
```

For _non-factor_ variables (`Participant` is numeric and `Face` is character/string), [complete](https://tidyr.tidyverse.org/reference/complete.html) finds all [unique](https://stat.ethz.ch/R-manual/R-devel/library/base/html/unique.html) values for each column and finds all combinations of these elements. However, if a variable is a [factor](#factors), [complete](https://tidyr.tidyverse.org/reference/complete.html) uses it [levels](https://stat.ethz.ch/R-manual/R-devel/library/base/html/levels.html), even if not all levels are present in the data. E.g., we can use `Face` as factor with three levels: "M-1", "M-2", and "F-1". In this case, information is missing for both participants and should be filled with NAs. This approach is useful if you know all combinations that _should_ be present in the data and need to ensure the completeness.
```{r}
extended_df <-
incomplete_df %>%
# converting Face to factor with THREE levels (only TWO are present in the data)
mutate(Face = factor(Face, levels = c("M-1", "M-2", "F-1"))) %>%

# completing the table
complete(Participant, Face)
```
```{r echo=FALSE}
knitr::kable(extended_df)
```

You can also supply default values via `fill` parameter that takes a [named](#names) [list](#lists), e.g., `list(column_name = default_value)`:
```{r}
filled_df <- incomplete_df %>%
  complete(Participant, Face, fill=list(Attractiveness=0, Symmetry=0))
```
```{r echo=FALSE}
knitr::kable(filled_df)
```

::: {.practice}
Do exercise 6.
:::





  ## Dropping / omtting NAs {#na.omit}

  There are two approaches on excluding missing values. You can exclude all incomplete rows which have missing values in _any_ variable via [na.omit()](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/na.fail.html) (base R function) or [drop_na()](https://tidyr.tidyverse.org/reference/drop_na.html) (_tidyr_ package function). Or you can exclude rows only if they have `NA` in a specific columns by specifying their names.

For a table you see below
```{r echo=FALSE}
widish_df_with_NA <-
  tibble(Participant = c(1, 1, 2, 2),
         Face = rep(c("M-1", "M-2"), 2),
         Symmetry = c(6, NA, 5, 3),
         Attractiveness = c(NA, 7, 2, 7),
         Trustworthiness = c(3, NA, 1, 2))

widish_df_with_NA %>%
  knitr::kable()
```

First, we can ensure only complete cases
```r
na.omit(widish_df_with_NA)
```
```{r echo=FALSE}
na.omit(widish_df_with_NA) %>%
  knitr::kable()
```

```r
widish_df_with_NA %>%
  drop_na()
```
```{r echo=FALSE}
widish_df_with_NA %>%
  drop_na() %>%
  knitr::kable()
```

Second, we drop rows only if `Attractiveness` data is missing.
```r
widish_df_with_NA %>%
  drop_na(Attractiveness)
```
```{r echo=FALSE}
widish_df_with_NA %>%
  drop_na(Attractiveness) %>%
  knitr::kable()
```

Practice time. Create you own table with missing values and exclude missing values using `na.omit()` and `na_drop()`.

::: {.practice}
Do exercise 7.
:::

  `na_drop` is a very convenient function but you can replicate it functionality using [is.na()](https://stat.ethz.ch/R-manual/R-devel/library/base/html/NA.html) in combination with [filter](#filter) dplyr function or logical indexing. Implement code that excludes rows if they contain `NA` in a specific column using these two approaches.

    ::: {.practice}
    Do exercise 8.
    :::


      Now, implement code that uses logical indexing as a [function](#seminar05) that takes `data.frame` (table) as a first argument and name a of a single column as a second, filters out rows with `NA` in that column and return the table back.

        ::: {.practice}
        Do exercise 9.
        :::

          As noted above, you can also impute values. The simplest strategy is to use either a fixed or an average (mean, median, etc.) value. _tidyr_ function that performs a simple substitution is [replace_na()](https://tidyr.tidyverse.org/reference/replace_na.html)^[There is also an inverse function [na_if()](https://dplyr.tidyverse.org/reference/na_if.html) that converts a specific value to an `NA`.] and, as a second parameter, it takes a named list of values `list(column_name = value_for_NA)`. For our toy table, we can replace missing `Attractiveness` and `Symmetry` values with some default value, e.g. `0` and `-1` (this is very arbitrary, just to demonstrate how it works!)
        ```{r}
        widish_df_with_NA %>%
          replace_na(list(Attractiveness = 0, Symmetry = -1)) %>%
          knitr::kable()
        ```

        Unfortunately, `replace_na()` works only with constant values and does not handle grouped tables very well^[At least I wasn't able to figure out how to do this.] So to replace an `NA` with a mean value of a grouped data, we need to combine some of our old knowledge with an [ifelse(conditon, value_if_true, value_if_false)](https://stat.ethz.ch/R-manual/R-devel/library/base/html/ifelse.html) function. Here is how it works to replace all negative values in a vector with a `0` and all non-negative with `1`:
```{r}
v <- c(-1, 3, 5, -2, 5)
ifelse(v < 0, 0, 1)
```
It works _per element_ and you can use original values themselves. Here is how to replace only negative values but keep the positive ones:
```{r}
v <- c(-1, 3, 5, -2, 5)
ifelse(v < 0, 0, v)
```
We, essentially, tell the function, "if the condition is false, use the original value". Now, your turn! Using the same vector and `ifelse()` function, replace negative values with a [maximal](https://stat.ethz.ch/R-manual/R-devel/library/base/html/Extremes.html) value of the vector. I know it is `5` but you cannot hardcode it, that's the point of the exercise!

                                                                                                                      ::: {.practice}
                                                                                                                    Do exercise 10.
                                                                                                                    :::

                                                                                                                      Now that you know how to use `ifelse()`, replacing `NA` with a mean will be (relatively) easy. Use [adaptation_with_na](data/adaptation_with_na.csv) table and replace missing information using participant-specific mean. This is the same table as above, so you need to compute `Psame` first. However, missing values in `Nsame` and `Ntotal` will result in `NA` for a corresponding `Psame`. Replace them with mean `Psame` _per participant_. This will require knowledge of additional details. First, you can compute [mean()](https://stat.ethz.ch/R-manual/R-devel/library/base/html/mean.html) ignoring any `NA` using `na.rm` parameter. Second, you will need to use [is.na()](https://stat.ethz.ch/R-manual/R-devel/library/base/html/NA.html) to detect values that you need to replace with the mean. Third, reminder, in order to apply some function per participant (stimulus, etc.), you need to first [group](https://dplyr.tidyverse.org/reference/group_by.html) your data. This entire computation should be implemented as a single pipeline.

                                                                                                                    This is how the original table with missing values looks like (arranged by `Prime` and `Probe`)
                                                                                                                    ```{r echo=FALSE}
                                                                                                                    adaptation_na <-
                                                                                                                      read_csv("data/adaptation_with_na.csv",
                                                                                                                               col_types = cols(Participant = col_character(),
                                                                                                                                                Prime = col_character(),
                                                                                                                                                Probe = col_character(),
                                                                                                                                                Nsame = col_double(),
                                                                                                                                                Ntotal = col_double()
                                                                                                                               )) %>%
                                                                                                                      mutate(Psame = Nsame / Ntotal)

                                                                                                                    adaptation_na %>%
                                                                                                                      arrange(Prime, Probe) %>%
                                                                                                                      head(4) %>%
                                                                                                                      knitr::kable()
                                                                                                                    ```

                                                                                                                    And here is the same table with imputed participant mean (I show it in a separate column just so you can see when it has been imputed)

```{r echo=FALSE}
                                                                                                                    adaptation_na %>%
                                                                                                                      group_by(Participant) %>%
                                                                                                                      mutate(Pavg = mean(Psame, na.rm=TRUE)) %>%
                                                                                                                      mutate(Psame = ifelse(is.na(Psame), mean(Psame, na.rm = TRUE), Psame)) %>%
                                                                                                                      arrange(Prime, Probe) %>%
                                                                                                                      head(4) %>%
                                                                                                                      knitr::kable()
```                                                                                                                    

::: {.practice}
Do exercise 11.
:::

                                                                                                                      Great work!
