---
title: "Final Assignment"
author: "YOUR NAME GOES HERE!"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Final assignment

↓↓↓ Any libraries you need to include go here. ↓↓↓

```{r library}
```

## Experiment
In the single task (condition _single_), one letter was presented in one of the four quadrants of the screen and observer responded about its identity (T or L). In the dual task (condition _dual_), two letters were presented in two different quadrants of the screen, but only the **first**  of them (column `Location1`) was cued and observer responded about its identity. Columns `Correct` (1 if correct, 0 otherwise) and `Response_Time` (in seconds) describe observers’ report on a given trial.

The research question was whether two hemispheres have independent control over attention. specifically, in the _dual_ condition, two letters could be presented within left or right hemifields, so both had to be processed by a single hemisphere (respectively, right or left, _unilateral_ layout). Alternatively, when two letters were presented in upper or lower hemifields, each letter would be represented and processed in one hemisphere (again, letter from the left hemifield in the right hemisphere and vice versa, _bilateral_ layout). If attention is a single brain-wide resource then same amount of attention will be allocated to the two target, irrespective of whether they are both represented in a single hemisphere or one in each (same performance and response times for _unilateral_ and _bilateral_ layouts). However, if two hemispheres have independent attention resource then presentation of two targets to two hemispheres (_bilateral_ layout) would mean that each hemisphere can spend all its attention on a single target. Conversely, presenting both targets to one hemisphere (_unilateral_ layout) would force it to split its attention resource on two targets, limiting it. In this case, we would expect performance to be better and response times shorter for _bilateral_ than for _unilateral_. Theoretically, if attention is truly a per-hemisphere resource, we would expect that performance and response time in the _bilateral_ layout to be as good as for a _single_ target condition, as in both cases a single hemisphere needs to process just a single target. 


## Location encoding
Location is coded in `Location1` (_single_ task and the first quadrant of the _dual_ task) and in `Location2` (second quadrant for the _dual_ task) column starting from upper right quadrant and going counterclockwise as follows:

1. upper right,
2. upper left,
3. bottom left,
4. bottom right.

## CSV-file format

* `Observer`: observer identity
* `Condition`: either "single" or "dual"
* `Location1`: quadrant of the only (_single_ task) or of the first letter (_dual_ task)
* `Location2`: location of the second letter (_dual_ task) or NA (_single_ task)
* `Correct`: 1 if respons was correct, 0 otherwise
* `Response_Time`: in seconds.

## Grading

### Preprocessing

1. Combine individual data files into a single table. You must use `map_dfr` for this.

```{r}
```

### Single task
#### For 1.7

1. Compute average observer’s performance, number of correct trials, total number of trials, and median response time (RT) for each quadrant, **irrespective** of condition. Store results in a new table (please use a sensible name). Single pipe.
```{r}

```

2. Plot the averages with quadrant label on x-axis and mean performance or RT on y-axis using box plots.
```{r single accuracy boxplot}
```

```{r single RT boxplot}
```

3. Test whether quadrant had a significant effect on performance (`glmer`, binomial family) / RT (`lmer`, linear mixed models) with quadrant as a fixed factor and observer as a random. Check importance of the quadrant for both models via drop-one approach.

```{r single statistical analysis for logit(performance)}

```

```{r single statistical analysis for RT}

```

#### For 1.3

4. Repeat the analysis for 1.7 but for the left vs. right hemifield (create the new variable that will encode this).

```{r left/ right hemisfield accuracy table}
```

```{r left/ right hemisfield accuracy stats}
```

```{r left/ right hemisfield accuracy plot}
```


```{r left/ right hemisfield RT table}
```

```{r left/ right hemisfield RT stats}
```

```{r left/ right hemisfield RT plot}
```

#### For 1.0
5. Same as #4 but for upper and lower hemifields.
```{r upper/ lower hemisfield accuracy table}
```

```{r upper/ lower hemisfield accuracy stats}
```

```{r upper/ lower hemisfield accuracy plot}
```

```{r upper/ lower hemisfield RT table}
```

```{r upper/ lower hemisfield RT stats}
```

```{r upper/ lower hemisfield RT plot}
```


### Dual task

#### For 1.7

1. Compute average observer’s performance, number of correct trial, total number of trials and median response times for the presentation layout (**unilateral** versus **bilateral**). Create new variable `Layout` for this first.

```{r averages by dual}
```

2. Plot the averages with layout labels on x-axis and mean performance or RT on y-axis using boxplots.
```{r dual accuracy plot}
```

```{r dual RT plot}
```

3. Compare performance (`glmer`, binomial family) and RT (`lmer`, linear mixed models) in unilateral versus bilateral tasks using generalized linear-mixed models and drop-one testing.
```{r statistics accuracy}
```

```{r statistics RT}
```

#### For 1.3

4. Compare performance in **unilateral** layout to that in the **single task** using generalized linear-mixed models. For this, compute a separate table for a single task performance (irrespective of the target location) with `Layout` column value `"single task"`. Next, bind by rows with the table you created during previous analysis. You will need to filter out `"bilateral"` layout, you can do it either before or after binding tables.
```{r Accuracy comparison unilateral}
```

5. Same but **bilateral** vs **single task**.
```{r performance comparison bilateral}
```

#### For 1.0:
6. Plot single task, unilateral, and bilateral performance (x-axis) and performance on y axis using boxplots. Use _forcats_ library to ensure that the order is `single`, `bilateral`, `unilateral`.
```{r plot performance all}
```


Compare performance for all three layouts in a single statistical model (glmer).
```{r}
```
