---
title: "Final Assignment"
author: "YOUR NAME GOES HERE!"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Final assignment

↓↓↓ Any libraries you need to include go here. ↓↓↓

```{r library}
```

## Experiment

In the single task (condition _single_), a single letter was presented in one of the four quadrants of the screen and observer responded about its identity (T or L). In the dual task (condition _dual_), two letters were presented in two different quadrants of the screen, but only the **first**  of them (column `Location1`) was cued and observer responded about its identity. Columns `Correct` (1 if correct, 0 otherwise) and `Response_Time` (in seconds) describe observers’ report on a given trial.

Location is coded in `Location1` (_single_ task and the first quadrant of the _dual_ task) and in `Location2` (second quadrant for the _dual_ task) column starting from upper right quadrant and going counterclockwise as follows:

1. upper right,
2. upper left,
3. bottom left,
4. bottom right.

## CSV-file format

* `Observer`: observer identity
* `Condition`: either "single" or "dual"
* `Location1`: quadrant of the only (_single_ task) or of the first letter (_dual_ task)
* `Location2`: location of the second letter (_dual_ task) or NA (_single_ task)
* `Correct`: 1 if respons was correct, 0 otherwise
* `Response_Time`: in seconds.

## Grading

### Preprocessing

1. Combine individual data files into a single table. You must use `map_dfr` for this.

```{r loading and joining data}
```

### Single task
#### For 1.7

1. Compute average observer’s performance, logit(performance) (you can use function `logit` from the _car_ package) and median response time (RT) for each quadrant, **irrespective** of condition and store results in a separate table.
```{r grouping data and basic calculations}

```

2. Plot the averages with quadrant label on x-axis and mean performance or RT on y-axis using box plots.
```{r single accuracy boxplot}
```

```{r single RT boxplot}
```


3. Test whether quadrant had a significant effect on logit(performance) / RT using linear mixed models (package _lmerTest_) with quadrant as a fixed factor and observer as a random. Check importance of the quadrant for both models via drop-one approach.

**Here and below** use logit(performance) when performing statistical comparisons!
```{r statistical analysis for logit(performance)}

```

```{r statistical analysis for RT}

```

#### For 1.3

4. Compute additional column of whether the target was on the left or on the right. Use linear-mixed models to assess the difference between targets presented in left and right hemifield (again, use logit(performance) when performing statistical tests and use drop-one approach). Plot averages using box plots, with hemifield labels on x-axis. Overlay individual observers as dots us _ggbeeswarm_ library.

```{r left/ right hemisfield accuracy table}
```

```{r left/ right hemisfield accuracy stats}
```

```{r left/ right hemisfield accuracy plot}
```


```{r left/ right hemisfield RT table}
```

```{r left/ right hemisfield RT stats}
```

```{r left/ right hemisfield RT plot}
```


5. Same as #4 but for upper and lower hemifields.
```{r upper/ lower hemisfield accuracy table}
```

```{r upper/ lower hemisfield accuracy stats}
```

```{r upper/ lower hemisfield accuracy plot}
```

```{r upper/ lower hemisfield RT table}
```

```{r upper/ lower hemisfield RT stats}
```

```{r upper/ lower hemisfield RT plot}
```

#### For 1.0

6. Instead of computing mean performance, compute number of correct and number of total trials per observer and quadrant. Use generalized linear models with binomial family and drop-one.

```{r single binomial statistics accuracy}
```


### Dual task

#### For 1.7

1. Compute average observer’s performance and RT for the presentation layout (**unilateral** versus **bilateral**).

```{r averages by dual}
```

2. Plot the averages with layout labels on x-axis and mean performance or RT on y-axis using boxplots.
```{r dual accuracy plot}
```

```{r dual RT plot}
```


3. Compare performance and RT in unilateral versus bilateral tasks using linear-mixed models and drop-one testing.
```{r statistics accuracy}
```

```{r statistics RT}
```

#### For 1.3

4. Compare performance in **unilateral** layout to that in the **single task** using linear-mixed models.
```{r Accuracy comparison unilateral}
```


5. Compare performance in **bilateral** layout to that in the **single task** using linear-mixed models.
```{r performance comparison bilateral}
```

6. Plot single task, unilateral, and bilateral performance (x-axis) and performance on y axis using boxplots. Use _forcats_ library to ensure that the order is `single`, `bilateral`, `unilateral`.
```{r compare performance all}
```


#### For 1.0:

7.  Instead of computing mean performance, compute number of correct and number of total trials per observer and quadrant. Use generalized linear models with binomial family and drop-one.

```{r dual binomial}
```




