---
title: "Seminar 02, Vectors and Scalars"
author: "Alexander (Sasha) Pastukhov"
date: "30 10 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercise 1. Assignment statement
Take a look at the code below. What would be the value of `x` at the end? Write down your answer **before** running code chunks!

**Answer:**
```{r exercise 01-1}
x <- 5
x <- 3
print(x)
```

**Answer:**
```{r exercise 01-2}
x <- 5
x <- x * x
print(x)
```

**Answer:**
```{r exercise 01-3}
x <- 5
x <- x / x
x <- 2
print(x)
```

## Exercise 2. Assigning values (vectors) to variables

Take a look at the code below. What would be the results of computation in each chunk? Write down your answer **before** running code chunks!

**Answer:**
```{r exercise 02-1}
x <- c(1, 3, 5)
y <- x
x <- c(3, 4, 1)
z <- y
z
```

**Answer:**
```{r exercise 02-2}
x <- c(1, 3, 5)
y <- x
x <- c(3, 4, 1)
z <- x
z
```
**Answer:**
```{r exercise 02-3}
x <- c(1, 3, 5)
y <- x
x <- c(3, 4, 1)
z <- x
z <- 2
z
```


## Exercise 3. Concatenating vectors.

Take a look at the code below. What would be the results of the computation? Write down your answer **before** running the code chunk!
**Answer:**

```{r exercise 03}
x <- c(1, 3, 5)
y <- c(2, 4, 2)
z <- c(3, 4, 1)

c(y, 6, z, x, c(10, 11), x)
```

## Exercise 4. Vector indexing.
What will be an output of each chunk? Write down your answers and then run chunks to check.

**Answer #1:**
```{r exercise 04-01}
x <- c(2, 13, 4, 25, 4, 122, -12, 78)

# 1
print(x[2])
```

**Answer #2:**
```{r exercise 04-02}
#2
x <- c(2, 13, 4, 25, 4, 122, -12, 78)
x[0] <- 5
print(x)
```

**Answer #3:**
```{r exercise 04-03}
#3
x <- c(2, 13, 4, 25, 4, 122, -12, 78)
print(x[10])
```

**Answer #4:**
```{r exercise 04-04}
#4
x <- c(2, 13, 4, 25, 4, 122, -12, 78)
x[8] <- 8
print(x)
```

**Answer #5:**
```{r exercise 04-05}
#4
x <- c(2, 13, 4, 25, 4, 122, -12, 78)
x[12] <- 12
x[0] <- 13
print(x)
```

## Exercise 5. Zero and out-of-range indexing
What will be an output of each chunk? Write down your answers and then run chunks to check.


**Answer #1:**
```{r exercise 05-01}
x <- c(2, 13, 4, 25, 4, 122, -12, 78)
print(x[0])
```

**Answer #2:**
```{r exercise 05-02}
x <- c(2, 13, 4, 25, 4, 122, -12, 78)
print(x[10])
```

**Answer #3:**
```{r exercise 05-03}
x <- c(2, 13, 4, 25, 4, 122, -12, 78)
print(x[10])
```

**Answer #4:**
```{r exercise 05-04}
x <- c(2, 13, 4, 25, 4, 122, -12, 78)
x[10] <- 100
print(x[9])
```
**Answer #5:**
```{r exercise 05-05}
x <- c(13, 4, 25, 4, 122, -12, 78, 2)
x[10] <- 100
print(x[8])
```

**Answer #6:**
```{r exercise 05-06}
x <- c(13, 4, 25, 4, 122, -12, 78, 2)
x[0] <- 100
print(x)
```

## Exercise #6
What will be an output of each chunk? Write down your answers and then run chunks to check.

**Answer #1:**
```{r exercise 06-01}
x <- c(1, 2, 3, 4, 5)

# 1
print(x[-4])
```

**Answer #2:**
```{r exercise 06-02}
x <- c(1, 2, 3, 4, 5)

# 1
print(x[-10])
```

**Answer #3:**
```{r exercise 06-03}
x <- c(1, 2, 3, 4, 5)
x[-2] <- 10
x[-3] <- 8

# 1
print(x)
```

## Exercise #7. Named vectors
What will be an output of each chunk? Write down your answers and then run chunks to check.
**Answer #1:**
```{r exercise 07-01}
point <- c("x"=20, "y"=10, "z"=1)
point[3] <- point["x"] + point[2]
print(point["z"])
```
**Answer #2:**
```{r exercise 07-02}
point <- c("x"=20, "y"=10, "z"=1)
names(point)[3] <- "Zzzzz..."
print(names(point))
```
**Answer #3:**
```{r exercise 07-03}
point <- c("x"=20, "y"=10, "z"=1)
names(point) <- c("c", "b", "a")
print(point["a"])
```

## Exercise #8. Colon operator and slicing.
What will be an output of each chunk? Write down your answers and then run chunks to check. Hint, first figure out and write down vector values. Next, do the same for the index vector. Then, see what values you get will get from that index. 

**Answer #1:**
```{r exercise 08-01}
x <- 10:18
print(x[4])
```

**Answer #2:**
```{r exercise 08-02}
x <- 20:10
print(x[3:2])
```

**Answer #3:**
```{r exercise 08-03}
x <- 10:20
print(x[-4:-1])
```

**Answer #4:**
```{r exercise 08-04}
x <- 10:20
print(x[seq(0, 12, by=2)])
```


## Exercise 9. Vector operations.

Take a look at the code below. What would be the results of the computation? Write down your answer **before** running the code chunk!
**Answer:**

```{r exercise 09}
x <- c(1, 3, 5)
y <- c(2, 4, 2)
z <- c(3, 4, 1)

x * y + z
```

## Exercise 10. Vector operations, mismatching lengths

Take a look at the code below. What would be the results of the computation? Write down your answer **before** running the code chunk!

**Answer #1:**
```{r exercise 10-01}
x <- c(1, 3, 5, 2)
y <- c(2, 4)
z <- c(3, 4, 1, 4, 7, 2, 6, 2)

x * y + z
```

**Answer #2:**
```{r exercise 10-02}
x <- c(1, 3, 5, 2)
y <- c() # Hint: you concatenate values into a vector but there are zero values listed. What kind of vector you will get?
z <- c(3, 4, 1, 4, 7, 2, 6, 2)

x * y + z
```


## Exercise 11. Let's see if R can suprising you!
Take a look at the code below. What would be the results of the computation? Write down your answer **before** running the code chunk!

**Answer #1:**
```{r exercise 11-01}
3[1]
```

**Answer #2:**
```{r exercise 11-02}
3[1][1]
```

**Answer: #3**
```{r exercise 11-03}
x <- 1:10
x[5][1][2]
```

**Answer: #4**
```{r exercise 11-04}
x <- 1:10
x[1:5][1[1]]
```

**Answer: #5**
```{r exercise 11-05}
x <- 1:10
x[1:5][2[1]]
```

**Answer: #6**
```{r exercise 11-06}
x <- 1:10
x[0][-2]
```

**Answer: #7**
```{r exercise 11-07}
5[1:5][-2]
```

**Answer: #8**
```{r exercise 11-08}
5[-1]
```

## Exercise 12
Three very hungry participants attempt to break world record for most hot dogs eaten in 10 minutes (currently, 74). After 8 minutes they eaten 45, 69, and 57 hot dogs. Compute how many hot dogs per minute must each of the eat to break the record?

```{r exercise 12}
# your code goes here
```

## Exercise 13
Sasha and Natalia have 6 and 12 chocolate cakes, respectively. Natalia realizes that she has way too many and gives Sasha one third of her cakes. How many chocolate cakes do they need to eat each before their diet starts in ten minutes? Used named vectors to make it more personal.

```{r exercise 13}
# your code goes here
```

## Exercise 14
Andrej and Boris drink tea. Andrej has a 0.5 liter tea cup and put 10 spoon of sugar. Boris has a 0.2 liter tea cup and used 3 spoons of sugar. Compute relative tea sweetness for each boy. Use named vectors.

```{r exercise 14}
# your code goes here
```

## Exercise 15
Space X made following number of launches in each year from 2010 till 2017: 2, 0, 2, 3, 6, 7, 9, 18. Of which failures were: 0, 0, 0, 0, 0, 1, 1, 0.  Compute number of **successful** launches.

```{r exercise 15}
# your code goes here
```

## Exercise 16
Using information from exercise 15, compute the total number of launches. You can use `sum()` function. Type "?sum()" in console.
```{r exercise 16}
# your code goes here
```

## Exercise 17
Using information from exercise 15, compute average success per year.
```{r exercise 17}
# your code goes here
```

## Exercise 18
Using information from exercise 15, compute average success over the entire period.
```{r exercise 18}
# your code goes here
```
