---
title: "Seminar 05 - functions"
author: "YOUR NAME GOES HERE"
date: "26 11 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercise 01
Take a look at the function definition below and fill in the answers
```{r exercise 01-definition, eval=FALSE}
cos <- function(angle_in_radians){
  ...magic...
  return(a_number)
}
```

* What is the name of the function? **Answer:**
* How many parameters does it have? **Answer:**
* How would you use it to compute a cosine of 0, 0.2, and 0.3?
```{r exercise 01}

```


## Exercise 02
 Write the definition of a function that double a value, you do not need to write any code inside or a return statement at the end of it.
```{r exercise 02-1}

```

Write the code that implements the "double-the-value" computation and test it by creating a variable with the same name as your parameter inside the function.
```{r exercise 02-2}

```

Combined code from 02-1 and 02-2, make sure you `return` the result  of computation. Test it by calling the function in the same chunk (you can do this only _after_ the function was defined!).
```{r exercise 02-3}

```

## Exercise 03
Write a function that would convert angle in _degrees_ to _radians_. The formula is $$rad = \frac{deg \cdot \pi}{180}$$.
```{r exercise 03}

```


