---
title: "Simulations and resampling"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# add libraries that you require here
```

## Exercise 01
Generate 20 samples from a normal distribution, choose mean and standard deviation yourself. Plot histogram and adding a red vertical line to indicate the true mean of the distribution and a blue one for the sample mean.

```{r exercise-01}
```

## Exercise 02
Use `sample()` to resample data from exercise 01 with replacement.

```{r exercise-02}
```

## Exercise 03
Use `replicate()` and `sample()` to compute mean for 1000 resampled values. Plot the distribution of _means_ and compute percentile 95% confidence inteval via `quantile()`.

```{r exercise-03}
```

## Exercise 04
Repeat exercise 03 but using `boot` library.

```{r exercise-04}
```

## Exercise 05
Using `boot` library, create a function that takes number of samples as the only parameter, draws data from a normal distribution, bootstraps mean 1000 times, computes percential 95% confidence interval and returns it as a tibble.
```{r exercise-05}
```

## Exercise 06
Use `group_modify()` and your function from exercise 05 to create a table with `n_samples` (go from 10 to 100), `lowerCI`, and `upperCI`
```{r exercise-06}
```

## Exercise 07
Use `geom_ribbon()` and `geom_hline()` to plot results of exercise 06
```{r exercise-07}
```

## Exercise 08
Extend exercise 06 to repeat computation of CI 100 times for each number of samples. Compute average lower and upper CI limits and plot them.
```{r exercise-08}
```

## Exercise 09
Modify exercise 08 to include average sample mean.
```{r exercise-09}
```

## Exercise 10
Simulate effect of number of samples and probability of success using binomial distribution.
```{r exercise-10}
```
