---
title: "Seminar 02, Vectors and Scalars"
author: "Alexander (Sasha) Pastukhov"
date: "30 10 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercise 1

Take a look at the code below. What would be the results of both computations? Write down your answer **before** running code chunks!

**Answer:**
```{r exercise 01}
x <- c(1, 3, 5)
y <- x
x <- c(3, 4, 1)
z <- y
z
```

**Answer:**
```{r exercise 01}
x <- c(1, 3, 5)
y <- x
x <- c(3, 4, 1)
z <- x
z
```


## Exercise 2

Take a look at the code below. What would be the results of the computation? Write down your answer **before** running the code chunk!
**Answer:**

```{r exercise 02}
x <- c(1, 3, 5)
y <- c(2, 4, 2)
z <- c(3, 4, 1)

c(y, 6, z, c(10, 11), x)
```

## Exercise 3

Take a look at the code below. What would be the results of the computation? Write down your answer **before** running the code chunk!
**Answer:**

```{r exercise 03}
x <- c(1, 3, 5)
y <- c(2, 4, 2)
z <- c(3, 4, 1)

x * y + z
```

## Exercise 4


## Exercise 5
What will be an output of each chunk? Write down your answers and then run chunks to check.
**Answer #1:**
```{r exercise 05-01}
x <- c(2, 13, 4, 25, 4, 122, -12, 78)

# 1
print(x[2])
```
**Answer #2:**
```{r exercise 05-02}
#2
x <- c(2, 13, 4, 25, 4, 122, -12, 78)
x[0] <- 5
print(x)
```
**Answer #3:**
```{r exercise 05-03}
#3
x <- c(2, 13, 4, 25, 4, 122, -12, 78)
print(x[10])
```

**Answer #4:**
```{r exercise 05-04}
#4
x <- c(2, 13, 4, 25, 4, 122, -12, 78)
x[8] <- 8
print(x)
```
**Answer #5:**
```{r exercise 05-05}
#4
x <- c(2, 13, 4, 25, 4, 122, -12, 78)
x[12] <- 12
x[0] <- 13
print(x)
```
## Exercise #7
What will be an output of each chunk? Write down your answers and then run chunks to check.

**Answer #1:**
```{r exercise 06-01}
x <- c(1, 2, 3, 4, 5)

# 1
print(x[-4])
```

**Answer #2:**
```{r exercise 06-02}
x <- c(1, 2, 3, 4, 5)

# 1
print(x[-10])
```

**Answer #3:**
```{r exercise 06-02}
x <- c(1, 2, 3, 4, 5)
x[-2] <- 10
x[-3] <- 8

# 1
print(x)
```
