---
  title: "Seminar 11"
output: html_document
---

## Exercise 01.
Function `rep()` is frequently used to create a vector of same values, e.g., all zeros, ones, or `NA`. Create a table (either `data.frame` or `tibble`) with three columns (names are up to you) and 10 rows. One should be initialized to ones, second one to a  repeating pattern `1, 2, 3, 4`, and the last one to `NA`.

```{r exercise-01}

```

## Exercise 02.
Use `expand_grid()` function to create table with combinations of `gender` (values `female`, `male`) and `handedness` (values `right`, `left`, `ambidextrous`). However, you need to have _twelve_ rows, so that each combination appears twice. You can do it in one `expand_grid()` call, possibly, with a help of `rep()` function. Then, use piping and mutate to add `index` column that goes from 1 to number of rows (e.g., index of each participant). Use function [n()](https://dplyr.tidyverse.org/reference/context.html) to get number of rows in the table you are working on. 

```{r exercise-02}

```

## Exercise 03.
Use a for loop to replicate functionality of `rep()` function. Start with an empty vector `target` (an empty vector as when you create it via `c()` function but do not provide any elements) and add (again, `c()` function!) values `1, 4, 2` to it four times. Thus, you end result should be `1, 4, 2, 1, 4, 2, 1, 4, 2, 1, 4, 2`.

```{r exercise-03}

```

## Exercise 04.
Use nested loop to replicate functionality of `expand_grid()` and results of exercise 02 (but excluding the `index` column, just `gender` and `handedness`). You can create two empty vectors, grow them in parallel and then use them to create a table. 

```{r exercise-04}

```

## Exercise 05.
Generate a random walk (length 10) starting at value zero with each next step drawn randomly from a normal distribution with mean of 0.5 and standard deviation of 2 (the function you are looking for is [rnorm()](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/Normal.html)). Think about the first value you initialize your vector to, how many iterations you need and what the range of values you require. You can append values to the vector either via `c()` function or directly using index (remember, if there was no value for a given index, R will expand vector to that length).

```{r exercise-05}

```

## Exercise 06.
Read and merge all `Persistence` folder csv-files into a single full table. Create a full table variable (I, typically, call it `results` or `reports`) and initialize it to an empty `data.frame()`. Next, you loop over file names (use function [dir_ls()](https://www.rdocumentation.org/packages/fs/versions/1.5.0/topics/dir_ls) for this), [read](#readr) one file at a time (don't forget to specify column types or will get a lot of warnings), and then use [bind_rows](https://dplyr.tidyverse.org/reference/bind.html) to combine the full table and the new table you loaded. Note that [bind_rows](https://dplyr.tidyverse.org/reference/bind.html) returns a _new_ table, so you need to assign it back to the original full table variable. Once you are done, your table should have 5232 rows and twelve columns.

```{r exercise-06}

```

## Exercise 07.
Use `map_dfr()` to load  and merge all persistence study files, replicating result of exercise 6.  Use`~` call notation and remember that `.` would then correspond to a single value from the vector of file names. Again, you should get a single table with twelve columns and 5232 rows.

```{r exercise-07}

```

## Exercise 08.
Program a function that takes a filename, loads the file and returns total number of trials/rows (if you forgot how to compute number of rows in a table, see [above](#forloop)). Once you have a function, use it with `map_dbl` and a vector of persistence filenames. You should get a vector of ten values. Now we can easily see that there was something wrong with one of the files and we must pay attention to the amount of data that we have.

```{r exercise-08}

```
