---
title: "Seminar 10"
output: html_document
---

## Exercise 01.
Generate and plot left side of the tree using `x = y - 0.3 * floor(y)` formula

```{r exercise-01}

```

## Exercise 02.
Generate right side of the tree by mirroring its left side.

```{r exercise-02}

```

## Exercise 03.
Bind left and right tree sides into a single table and plot it.

```{r exercise-03}

```

## Exercise 04.
Plot left and right side using separate `geom_path()` calls.
```{r exercise-04}

```

## Exercise 05
Join left and right side of the tree into a new table and plot the tree using `geom_ribbon`
```{r exercise-05}

```

## Exercise 06.
Use a theme to remove the grid and add the background image.

```{r exercise-06}

```

## Exercise 07.
Sample `tree` table (created in exercise #5) and plot individual "balls" using `geom_point`. Use x-coordinates for the left edge for a moment.

```{r exercise-07}

```

## Exercise 08.
Generate a random x-position between left and right edges of the tree, plot the decoration.
```{r exercise-08}

```

## Exercise 09.
Generate ten normally distributed numbers, loop over them and print each one. However, `break` after the _fifth_ value. For this, you need to loop over indexes of values.
```{r exercise-09}

```

## Exercise 10.
Randomize size and color of the balls.
```{r exercise-10}

```

## Exercise 11.
Generate and plot a star (pick the number of vertices that you like) assuming size (radius) of 1
```{r exercise-11}

```

## Exercise 12.
Resize and move the star putting it on top of the Christmas tree.

```{r exercise-12}

```
