---
title: "Seminar 05 - functions"
author: "YOUR NAME GOES HERE"
date: "26 11 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercise 01
Take a look at the function definition below and fill in the answers
```{r exercise 01-definition, eval=FALSE}
cos <- function(angle_in_radians){
  ...magic...
  return(a_number)
}
```

* What is the name of the function? **Answer:**
* How many parameters does it have? **Answer:**
* How would you use it to compute a cosine of 0, 0.2, and 0.3?
```{r exercise 01}

```


## Exercise 02
 Write the definition of a function that double a value, you do not need to write any code inside or a return statement at the end of it.
```{r exercise 02-1}

```

Write the code that implements the "double-the-value" computation and test it by creating a variable with the same name as your parameter inside the function.
```{r exercise 02-2}

```

Combined code from 02-1 and 02-2, make sure you `return` the result  of computation. Test it by calling the function in the same chunk (you can do this only _after_ the function was defined!).

Store results in a local variable and return its value.
```{r exercise 02-3}

```


Return computation directly (no extra variable).
```{r exercise 02-4}

```

## Exercise 03
Write a function that would convert angle in _degrees_ to _radians_. The formula is $$rad = \frac{deg \cdot \pi}{180}$$.
```{r exercise 03}

```

## Exercise 04
Study the code and decide on what will be printed. Hint, it might help to write comments next to each line that uses a variable to state whether it is a global variable, parameter (can have the same name as a global variable, "overwriting" it), or a local variable (can have the same name as a global variable or a parameter, "overwriting" them). Once you wrote down the answer, run the code to check it.

**Answer #1:**
```{r exercise 04-1}
x <- 1
my_function <- function(){
  x <- x + 1
}
print(x)
```
**Answer #2:**
```{r exercise 04-2}
x <- 1
my_function <- function(){
  x <- x + 1
}
my_function()
print(x)
```

**Answer #3:**
```{r exercise 04-3}
x <- 1
my_function <- function(){
  x <- x + 1
  return(x)
}
x <- my_function()
print(x)
```
**Answer #4:**
```{r exercise 04-4}
x <- 1
my_function <- function(x){
  y <- x + 1
  return(y)
}
x <- my_function(5)
print(x)
```

**Answer #5:**
```{r exercise 04-5}
x <- 1
my_function <- function(x){
  x <- 2
  return(x)
}
x <- my_function(5)
print(x)
```

**Answer #6:**
```{r exercise 04-6}
x <- 1
my_function <- function(x){
  y <- x + 1
  return(x)
}
x <- my_function(5)
print(x)
```
**Answer #7:**
```{r exercise 04-7}
x <- 1
my_function <- function(x){
  y <- x + 1
  return(x  + y)
}
x <- my_function(x) + x
print(x)
```
## Exercise 05
Write a function that takes _three_ parameters rather than one, add the first two, subtracts the third one, and returns the results of the computation.
```{r exercise 05}

```


## Exercise 06
use [mpg](https://ggplot2.tidyverse.org/reference/mpg.html) table from the _ggplot2_ package. Write a function that takes a table as a _parameter_ (so it does not assume that table exists in the global environment) and computes and returns average miles-per-gallon efficiency based on city `cty` and highway `hwy` test cycles. 

Compute and return a vector of values based on the table passed as parameter. 
```{r exercise 06-1}
```

Do the same computation but add the result _to the table_ as `mpg` column and return the entire table.
```{r exercise 06-2}
```

## Exercise 07
Write function that computes [mean](https://stat.ethz.ch/R-manual/R-devel/library/base/html/mean.html) efficiency for a particular cycle, either city or highway. For this, the function will take _two_ parameters: the table itself and a string with the name of the column. You can then use it to access the column via [double square brackets](#get-single-column). In short, you function takes a table and a string with a column name and returns a single number (mean for the specified column).
```{r exercise 07}
```

