## Looking at similarity
A different study, which used same four objects, showed that a similar looking history effect but for longer interruptions (1000 ms rather than 50 ms) was modulated by objects similarity. Let us check that hypothesis by computing a rough difference measure. It will assume that their difference is proportional to the absolute "distance" between them on x-axis in the above plot^[This measure assumes metric distance, which is a very strong assumption.]. E.g., distance between a sphere and a sphere is 0, but between sphere and quad-band or single-band and dual-band is 1. Difference between sphere and dual-band is 2, etc. You can compute it by converting _factor_ variables `Prime` and `Probe` to integers (this assumes that levels are in the correct order). Then, you can compute the [absolute difference](https://stat.ethz.ch/R-manual/R-devel/library/base/html/MathFun.html) between those indexes and store it as a new column (e.g. `Difference`). Next, group by `Difference` and `Participant` to compute average probability of the same response. Your plot should look like this (you will need to map `Difference` on `group` to get four box plots rather than one).

```{r echo=FALSE, fig.align='center'}
results |>
  select(-filename) |>
  mutate(SameResponse = Response1 == Response2)  |>
  mutate(Prime = factor(Prime, 
                        levels= c("heavy poles sphere", "stripes-8", "stripes-4", "stripes-2"),
                        labels = c("sphere", "quad band", "dual band", "single band")),
         Probe = factor(Probe, 
                        levels= c("heavy poles sphere", "stripes-8", "stripes-4", "stripes-2"),
                        labels = c("sphere", "quad band", "dual band", "single band")),) |>
  group_by(Participant, Prime, Probe) |>
  summarise(Psame = mean(SameResponse), .groups="drop") |>
  
  ungroup() |>
  mutate(iPrime = as.integer(Prime),
         iProbe = as.integer(Probe),
         Difference = abs(iPrime - iProbe)) |>
  
  group_by(Participant, Difference) |>
  summarise(Pavg = mean(Psame), 
            CIlower = quantile(Psame, (1-0.89)/2),
            CIupper = quantile(Psame, 1-(1-0.89)/2),
            .groups="drop") |>
  
  ggplot(aes(x=Difference, y=Pavg, ymin=CIlower, ymax=CIupper, group=Difference)) + 
  geom_boxplot()
```

::: {.infobox .practice}
Do exercise 8.
:::


## Exercise 08. Difference measure.
Compute difference between shapes by converting _factor_ variables `Prime` and `Probe` to integers (this assumes that levels are in the correct order). Then, you can compute the [absolute difference](https://stat.ethz.ch/R-manual/R-devel/library/base/html/MathFun.html) between those indexes and store it as a new column (e.g. `Difference`). Next, group by `Difference` and `Participant` to compute average probability of the same response. Your plot should look like this (you will need to map `Difference` on `group` to get four box plots rather than one).

```{r exercise 08}
```
