---
title: "Seminar 11"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# add libraries that you require here
```

## Exercise 01
Generate random data from normal distribution (see text for details). Compute Pearson's correlation coefficient using the formula in text. Compare it to the output of cor() function. You do not need to use loops to compute Pearson's correlation coefficient by hand.

```{r exercise-01}
```


## Exercise 02
Compute 95% confidence interval by resampling _pairs_ of `x` and `y` data that you generated in exercise 01. You can sample indexes in combination with replicate() and quantile() or use boot() and boot.ci(). Compare your results with the output of cor.test()

```{r exercise-02}
```

## Exercise 03
Perform a simulation for how standard deviation of noise affects confidence interval width. Generate 1000 random data sets for each level of noise. Plot the results as in the text.

```{r exercise-03}
```


## Exercise 04
Perform a simulation for how number of samples affects confidence interval width. Generate 1000 random data sets for each number of samples. Plot the results as in the text.

```{r exercise-04}
```

## Exercise 05
Generate data as in the text, compute the t value using the formula. Compare it to the output of t.test() function (should be the same but for rounding)

```{r exercise-05}
```

## Exercise 06
Use pairwise resampling of data to compute 95% percentile confidence inteval. Again, you can sample() index of pairs in combination with replicate() and quantile() or use boot() + boot.ci(). Compare your bootstrapped 95% CI with that reported by t.test()

```{r exercise-06}
```

## Exercise 07
Compute t value while randomly permuting values. Compute the proportion of t values that are smaller than the t value for the original data. Compare that to the output of t.test(..., alternative  = "less").

```{r exercise-07}
```


## Exercise 08
Generated linear model data as shown in the text. Use resampling of rows (slice_sample()) to generated samples, fit a model, and extract estimates by broom::tidy(). Resample 1000 times and compute 95% confidence interval for each term. Compare it with an output of summary(lm(...)) for the original data.

```{r exercise-08}
```

## Exercise 09
Compute p value for term x but permuting it (shuffling the order of values) 1000 times, fitting the model to permuted data and extracting t value. Compute proportion of absolute values of permuted t values that exceed the absolute t value for the original (unpermuted) data. Compare this to a p value for `x` in summary(lm(...)) output.

```{r exercise-09}
```

