---
  title: "Seminar 09"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

## Exercise 01.
Analyze _GP.csv_ data, a questionnaire on gaming habits, which was conducted prior to an experiment to check whether two groups of participants assigned to _Game_ and _Experiment_ conditions have similar gaming habits. We would like to visually inspect responses to individual items in a questionnaire appear for different conditions, as this will tell us whether we should expect a difference. Split the computations below into two pipelines. One that loads and pre-processes the data (steps 1-3). Another one that produces a summary and stores it into a different table (step 4). Advice, implement it one step at a time, checking the table and making sure that you get expected results before piping it and adding the next operation.

1. Read the file, make sure you specify column types.
2. Convert `Condition` column to a factor with `1` corresponding to `"Game"` and `2` to `"Exp"`.
3. Pivot all `GP..` columns. You should get a table with five columns: `Respondent`, `Condition`, `Gender`, `name` (or a column name that you specified), and `value` (or a column name that you specified). Hint, you can use slicing `:` to specify the range of columns or [starts_with()](https://tidyselect.r-lib.org/reference/starts_with.html) function to specify a common prefix. Try both approaches.
4. Group data by condition and GP item and compute median and [median absolute deviation](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/mad.html) of responses. These are robust versions of mean and standard deviation, better suitable for data with potential outliers.

Now you have a table that has median and MAD values for each combination and item. Plot them to compare them visually. Use median responses for y-value of points and median±MAD for error bars. Use facets and color to make it easier to identify the items and conditions.
```{r exercise 01}

```

## Exercise 02
Perform similar analysis but do not group data and summarize the data. Instead, use box plots to show the variability.

```{r exercise 02}

```

## Exercise 03

Pivot _adaptation_ data to a wide format that is easier for humans to read.

1. Read the file (don't forget to specify column types)
2. Computer `Psame` proportion of same responses given number of total responses.
3. Pivot the table wider, think about your id columns. Also try without specifying any and see what you get.
4. Compute an average stability across all probes and put it into a new `Average` column. You can do it "by hand" or you can use [rowSums()](https://stat.ethz.ch/R-manual/R-devel/library/base/html/colSums.html) to compute it. In the latter case, you need to use `.` to refer to the table inside the `mutate()` function and you will need to normalize it by the number of probes to get an average instead of the sum.
5. Pipe it to the output, using [knitr::kable()](https://bookdown.org/yihui/rmarkdown-cookbook/kable.html).

```{r exercise 03}

```

## Exercise 04
Using _adaptation_ data, create group average summary as a square 5×4 table with a single row per _Prime_ and four columns for _Probe_ plus a column that says which prime the row corresponds to. As a value for each cell, we want to code _median ± MAD_.
```{r exercise 04}

```

## Exercise 05
Using _persistence_ data, extract `Hour` and `Minutes` from `Session` column. Do it in a single pipeline, starting with reading the file. 
```{r exercise 05}

```

## Exercise 06
Using same [persistence](data/persistence.csv) data extract birth year and gender of participants from `Participant` code (keep the code though!). Put a nice extra touch by converting year to a number and gender into a factor type.

```{r exercise 06}

```


## Exercise 07
Create you own table with missing values and exclude missing values using `na.omit()` and `na_drop()`.

```{r exercise 07}

```

## Exercise 08
Replicate functionality of `na_drop()` using [is.na()](https://stat.ethz.ch/R-manual/R-devel/library/base/html/NA.html) in combination with [filter](#filter) dplyr function or logical indexing. Implement code that excludes rows if they contain `NA` in a specific column using these two approaches.

```{r exercise 08}

```

## Exercise 09
Implement code from the previous exercise that used logical indexing as a [function](#seminar05) that takes `data.frame` (table) as a first argument and name a of a single column as a second, filters out rows with `NA` in that column and return the table back.

```{r exercise 09}

```

## Exercise 10
Use  `ifelse()` function to replace negative values with a [maximal](https://stat.ethz.ch/R-manual/R-devel/library/base/html/Extremes.html) value of the vector. I know it is `5` but you cannot hardcode it, that's the point of the exercise!

```{r exercise 10}
v <- c(-1, 3, 5, -2, 5)
```

## Exercise 11
Use _adaptation_with_na_ table and replace missing information using participant-specific mean. This is the same table as above, so you need to compute `Psame` first. However, missing values in `Nsame` and `Ntotal` will result in `NA` for a corresponding `Psame`. Replace them with mean `Psame` _per participant_. This will require knowledge of additional details. First, you can compute [mean()](https://stat.ethz.ch/R-manual/R-devel/library/base/html/mean.html) ignoring any `NA` using `na.rm` parameter. Second, you will need to use [is.na()](https://stat.ethz.ch/R-manual/R-devel/library/base/html/NA.html) to detect values that you need to replace with the mean. Third, reminder, in order to apply some function per participant (stimulus, etc.), you need to first [group](https://dplyr.tidyverse.org/reference/group_by.html) your data. This entire computation should be implemented as a single pipeline.

```{r exercise 11}
```

